---
title: "Functions and Data Handling in R"
output:
  html_notebook:
    toc: true
    toc_depth: 3
---

## Learning Objectives

After completing this lab, relational will be able to:

- Create and use variables in R
- Write simple user-defined functions
- Read data from CSV files
- Explore and inspect datasets
- Create new variables from existing data
- Save modified datasets to files

## 1. Creating Variables

Variables store data values.

**Numeric Variables**
```{r}
height <- 5.7
name <- "Wakar"
height 
name 
```
**Character Variables**
```{r}
adress <- "Bashundhara"
university <- "AIUB" 
adress 
university
```

**Logical Variables**
```{r}
reached <- TRUE
!reached  
```

## 2. Working with Vectors
A vector stores multiple values of the same type.

```{r}
id <- c(5, 7, 9, 8, 6)
id
```
Access elements:
```{r}
id[5]
id[4]
```

Basic operations:

```{r}
mean(id)
max(id)
min(id)
sum(id)
```

## 3. Writing Simple Functions
Functions help reuse code.

General syntax:

function_name <- function(parameters){

    statements

    return(value)

}

**Example 1: Addition Function**
```{r}
add_numbers <- function(a, b){ 
  result <- a + b 
  return(result) 
} 

add_numbers(1, 90)


add_numbers(5,890)
```

**Example 2: Area of Rectangle**
```{r}
rectangle_area <- function(length, width){
  area <- length * width
  return(area)
}

rectangle_area(10, 5)
```

**Example 3: Grade Function**
```{r}
calculate_grade <- function(score){

  if(score >= 80){
    return("A")
  }

  else if(score >= 70){
    return("B")
  }

  else if(score >= 60){
    return("C")
  }

  else{
    return("F")
  }

}

calculate_grade(85)
```

## 4. Built-in Dataset Exploration

R provides several built-in datasets.

Load the iris dataset:

`data(iris)`

Display first rows:

`head(iris)`

Display last rows:

`tail(iris)`

## 5. Basic Dataset Inspection
Structure of Dataset
`str(iris)`

Shows:

- Variable names
- Data types
- Sample values

**Dataset Dimensions**
`dim(iris)`

Output:

`150 rows and 5 columns`

**Column Names**

`names(iris)`


**Summary Statistics**
`summary(iris)`

**View Dataset**

`View(iris)`

## 6. Reading Data from Files

**Reading CSV Files**

relational.csv

Example:

| ID | Name   | CGPA |
|----|--------|------|
| 1  | Rahim  | 3.70 |
| 2  | Karim  | 3.50 |
| 3  | Nabila | 3.90 |

**Read CSV File**

`relational <- read.csv("relational.csv")`

Alternative (Recommended)
```{r}

library(readr)
relational <- read_csv("relational.csv")
```

Display data:

`head(relational)`

**Check Structure**

`str(realational)`

**Reading Text Files**

Example: relational.txt

ID Name CGPA

1 Rahim 3.70

2 Karim 3.50

3 Nabila 3.90

```r
#Read Text File
relational <- read.table(
    "relational.txt",
    header = TRUE
)

head(relational)
```

**Read Tab-Separated Text File**

Example:

ID	Name	CGPA

1	Rahim	3.70

2	Karim	3.50

3	Nabila	3.90
```r
relational <- read.delim(
    "relational.txt"
)

head(relational)
```

**Reading Excel Files (.xlsx)**

```{r}
#Install package:
install.packages("readxl")

#Load package:
library(readxl)

#Read Excel file:

relational <- read_excel(
    "relational.xlsx"
)

head(relational)
```

Read Specific Worksheet

`relational <- read_excel(
    "relational.xlsx",
    sheet = "Sheet1"
)`


**Reading Data from URLs**

Many datasets are available online.

Example:
```{r}

url <- "https://raw.githubusercontent.com/mwaskom/seaborn-data/master/iris.csv"

iris_data <- read.csv(url)

head(iris_data)
```

**Reading Data from GitHub**

GitHub hosts many publicly available datasets.
```{r}
github_url <- "https://raw.githubusercontent.com/cs109/2014_data/master/countries.csv"

countries <- read.csv(
    github_url
)

head(countries)
```

**Reading Data from Google Drive**

Read a Public CSV Directly from Google Drive

Suppose the Google Drive sharing link is:

https://drive.google.com/file/d/FILE_ID/view?usp=sharing

Example:

https://drive.google.com/file/d/1ABCDEF123456789/view?usp=sharin

<span style="color:red;">
The file must be shared publicly or accessible to your account.
</span>

The File ID is: 
1ABCDEF123456789

Create a direct download URL:
```{r}
file_id <- "1ABCDEF123456789"

url <- paste0(
  "https://drive.google.com/uc?export=download&id=",
  file_id
)

#if the file is CSV
relational <- read.csv(url) 

head(relational)
```

## 7. Creating New Variables

Suppose the dataset contains:

relational

Create a scholarship variable:

`relational$Scholarship <- relational$CGPA >= 3.75`

Display:

`relational`

**Create Grade Category**
```{r}
relational$Grade <- ifelse(
    relational$CGPA >= 3.75,
    "Excellent",
    "Good"
)
```

Display:

`relational`



## 8. Writing Data into Files

Save dataset as CSV.
```{r}
write.csv(
    relational,
    "relational_updated.csv",
    row.names = FALSE
)
```

A new file named:relational_updated.csv will be created.

Since this is only the **second lab** of an undergraduate Introduction to Data Science course, the exercises should gradually move from **variables → functions → reading data → data inspection → creating variables → saving data**.

I would organize them into three levels: Basic, Intermediate, and Challenge.


## Lab Exercises

### Exercise 1: Creating Variables

Create the following variables:

* Student Name
* Student ID
* Age
* CGPA

Display all variables.

Expected output:

```{r} 

Name <- "Wakar"
ID<- 23-51784-2
Age<- 25
CGPA<- 3.47
Name
ID
Age
CGPA
```

### Exercise 2: Working with Vectors

Create a vector containing the following quiz marks:

```{r}


Total_marks = 100
Average_marks <- 70
Highest_mark <- 80
Lowest_mark <-60

Total_marks
Average_marks
Highest_mark
Lowest_mark


```
### Exercise 3: Creating Functions

Write a function that calculates:

```text
Area = Length × Width
```

Test the function using:

```text
Length = 12
Width = 8
```


### Exercise 4: Temperature Conversion

Write a function that converts Celsius to Fahrenheit.

Formula:

```text
F = (9/5 × C) + 32
```

Test using:

```text
C = 25
```

### Exercise 5: Exploring the mtcars Dataset

Load the built-in dataset:

```r
data(mtcars)
```

Perform the following tasks:

1. Display the first 10 rows.
2. Display the last 10 rows.
3. Determine the number of rows and columns.
4. Display column names.
5. Display summary statistics.

### Exercise 6: Dataset Inspection

Using the mtcars dataset:

Display:

```r
head(mtcars)
tail(mtcars)
str(mtcars)
summary(mtcars)
```

Briefly explain what each function does.

### Exercise 7: Reading a CSV File

Suppose the file:

```text
relational.csv
```

contains:

| ID | Name   | CGPA |
| -- | ------ | ---- |
| 1  | Rahim  | 3.70 |
| 2  | Karim  | 3.50 |
| 3  | Nabila | 3.90 |

Read the file and:

1. Display the first rows.
2. Display its structure.
3. Display summary statistics.

### Exercise 8: Reading an Excel File

Read:

```text
relational.xlsx
```

Display:

```r
head(data)
summary(data)
```

### Exercise 9: Reading Data from an Online Source

Load the Iris dataset directly from:

```r
url <- "https://raw.githubusercontent.com/mwaskom/seaborn-data/master/iris.csv"
```

Display:

* First 5 rows
* Dataset dimensions
* Summary statistics


### Exercise 10: Scholarship Eligibility

Using the following dataset:

| ID | Name   | CGPA |
| -- | ------ | ---- |
| 1  | Rahim  | 3.80 |
| 2  | Karim  | 3.40 |
| 3  | Nabila | 3.95 |

Create a new variable:

```text
Scholarship
```

Rule:

```text
CGPA ≥ 3.75 → TRUE
Otherwise → FALSE
```

### Exercise 11: Performance Category

Create a variable:

```text
Performance
```

Rule:

```text
CGPA ≥ 3.75      → Excellent
3.50–3.74        → Good
Below 3.50       → Average
```

Display the updated dataset.

### Exercise 12: Save a Modified Dataset

After creating the Scholarship and Performance variables:

Save the dataset as:

```text
relational_updated.csv
```

Verify that the file is created successfully.


## Mini Project

Create a dataset for at least 10 relational containing:

* ID
* Name
* Quiz Marks
* Assignment Marks
* Midterm Marks

Perform the following:

1. Read the dataset.
2. Inspect the dataset.
3. Create a Total Marks variable.
4. Create a Grade variable.
5. Display summary statistics.
6. Save the updated dataset as:

```text
final_results.csv


```
