---
title: "Statistical Thinking and Understanding Data"
output:
  html_notebook:
    toc: true
    toc_depth: 3
  html_document:
    toc: true
    toc_depth: '3'
    df_print: paged
---

## Learning Objectives

After completing this lab, students will be able to:

- Import datasets into R
- Identify variable types
- Generate summary statistics
- Calculate mean, median, variance, and standard deviation
- Convert data types

## 1. Importing Datasets

**Load a Built-in Dataset**

```{r}
data(mtcars)

head(mtcars)
```
**Dataset Dimensions**
```{r}
dim(mtcars)
```

Output: 

`32 11`

Meaning:

- 32 observations
- 11 variables

**View Dataset Structure**
```{r}
str(mtcars)
```

Output:

```r
'data.frame':	32 obs. of  11 variables:
 $ mpg : num  21 21 22.8 21.4 18.7 18.1 14.3 24.4 22.8 19.2 ...
 $ cyl : num  6 6 4 6 8 6 8 4 4 6 ...
 $ disp: num  160 160 108 258 360 ...
 $ hp  : num  110 110 93 110 175 105 245 62 95 123 ...
 $ drat: num  3.9 3.9 3.85 3.08 3.15 2.76 3.21 3.69 3.92 3.92 ...
 $ wt  : num  2.62 2.88 2.32 3.21 3.44 ...
 $ qsec: num  16.5 17 18.6 19.4 17 ...
 $ vs  : num  0 0 1 1 0 1 0 1 1 1 ...
 $ am  : num  1 1 1 0 0 0 0 0 0 0 ...
 $ gear: num  4 4 4 3 3 3 3 4 4 4 ...
 $ carb: num  4 4 1 1 2 1 4 2 2 4 ...
```
## 2. Identifying Variable Types
**Display Data Types**
```{r}
sapply(mtcars, class)
```

Output:

```r
 mpg       cyl      disp        hp 
"numeric" "numeric" "numeric" "numeric" 
     drat        wt      qsec        vs 
"numeric" "numeric" "numeric" "numeric" 
       am      gear      carb 
"numeric" "numeric" "numeric"
```

**Understanding Variable Types**

| Variable     | Type                 |
| ------------ | -------------------- |
| mpg          | Numeric              |
| cyl          | Numeric              |
| disp         | Numeric              |
| hp           | Numeric              |
| drat         | Numeric              |

**Count Variable Types**
```{r}
table(sapply(mtcars, class))
```

## 3. Summary Statistics

**Summary of Entire Dataset**
```{r}
summary(mtcars)
```

Output includes:

- Min
- 1st Quartile
- Median
- Mean
- 3rd Quartile
- Max

**Summary of a Single Variable**
```{r}
summary(mtcars$hp)
```

## 4. Measures of Central Tendency
**Mean**
```{r}
mean(mtcars$hp)
```

## Median 
```{r}
median(mtcars$hp)

```



**Mode (Custom Function)**
```{r}
mode(mtcars$hp)
```



```{r}
get_mode <- function(x)
{
  unique_values <- unique(x)

  unique_values[
    which.max(tabulate(match(x, unique_values)))
  ]
}

get_mode(mtcars$hp)
```

## 5. Measures of Spread
**Variance**
```{r}
var(mtcars$hp)
```

**Standard Deviation**
```{r}
sd(mtcars$hp)
```

**Range**
```{r}
range(mtcars$hp)
```

**Interquartile Range (IQR)**
```{r}
IQR(mtcars$hp)
```

## 6. Calculating Statistics for Multiple Variables

**Mean of All Numeric Variables**
```{r}
sapply(mtcars[,1:11], mean)
```

**Median of All Numeric Variables**
```{r}
sapply(mtcars[,1:11], median)
```

**Standard Deviation of All Numeric Variables**
```{r}
sapply(mtcars[,1:11], sd)
```

## 7. Converting Data Types
Create a sample dataset:

```{r}
Participants <- data.frame(
  ID = c("1001","1002","1003"),
  Age = c("25","26","27"),
  CGPA = c("3.00","3.50","3.80")
)

Participants
```

**Check Current Types**
```{r}
str(Participants)
```

Output:

```r
ID   : chr
Age  : chr
CGPA : chr
```

**Convert Character to Numeric**

```{r}
Participants$ID <- as.numeric(Participants$ID)

Participants$CGPA <- as.numeric(Participants$CGPA)
```

**Verify Conversion**
```{r}
str(faculty)
```

Output:

```r
Age  : num
CGPA : num
```

**Convert Character to Factor**

```{r}
Participants$ID <- as.factor(Participants$ID)

str(Participants)
```

## 8. Importing Data from CSV
Create a CSV file named:
`Book2.csv  `

Contents:

| ID | Name   | CGPA |
| -- | ------ | ---- |
| 1  | Rahim  | 3.70 |
| 2  | Karim  | 3.50 |
| 3  | Nabila | 3.90 |


Read file:

```{r}
faculty <- read.csv(
  "Book1.csv"
)

head(faculty)
```

## Exercises
**Exercise 1**

Load the built-in dataset:

```{r}
data(mtcars)

head(mtcars)
```
Find:

- Number of rows
- Number of columns
- Variable names

**Exercise 2**

For the variable:

`mtcars$mpg`

Calculate:

- Mean
- Median
- Variance
- Standard Deviation


**Exercise 3**

Display summary statistics for:

`mtcars`

**Exercise 4**

Create:

`ages <- c("18","20","22","25")`

Convert it to numeric and verify the type.

**Exercise 5**

Import a CSV file and determine:

- Dataset dimensions
- Variable types
- Summary statistics

## Important Points
- When calculating individual descriptive statistics, always remember to include na.rm = TRUE. If your column contains even a single missing value (NA), R will output NA as the result unless you explicitly tell it to ignore missing values.

- If you are  asked to calculate these metrics across different groups (e.g., finding the mean salary split by department). The **dplyr** package handles this beautifully.

**Example:**

```{r}
library(dplyr)
grouped_stats <- my_data_csv %>%
  group_by(categorical_column) %>%
  summarise(
    Group_Mean     = mean(numeric_column, na.rm = TRUE),
    Group_Median   = median(numeric_column, na.rm = TRUE),
    Group_Variance = var(numeric_column, na.rm = TRUE),
    Group_SD       = sd(numeric_column, na.rm = TRUE),
    Sample_Count   = n() # Counts number of observations per group
  )

print(grouped_stats)
```

## Homework

Download any public CSV dataset from Kaggle, UCI Machine Learning Repository, or another source.

Perform the following:

- Import the dataset into R.
- Display the first 10 rows.
- Display dataset dimensions.
- Identify variable types.
- Generate summary statistics.
- Calculate the mean, median, variance, and standard deviation for at least one numerical variable.
- Convert at least one variable to another data type.
Write a short summary (3–5 sentences) describing the dataset.
