---
title: "EDA - Visualization"
output:
  html_notebook:
    toc: true
    toc_depth: 3
---

## Learning Objectives

After completing this lab, students will be able to:

- Create histograms and boxplots
- Generate scatterplots using ggplot2
- Identify outliers visually
- Perform correlation analysis
- Prepare a simple EDA report

## 1. Load Required Libraries


```{r}
# Install packages (run once)
# install.packages("tidyverse")
# install.packages("corrplot")

library(tidyverse)
library(corrplot)
```
tidyverse is a collection of R packages designed for data science that provides a consistent and user-friendly framework for data import, cleaning, wrangling, visualization, and analysis. It simplifies common data manipulation tasks and promotes the use of tidy data, where each variable is a column, each observation is a row, and each value occupies a single cell.

Some commonly used packages in the tidyverse include:

- readr – Import data
- dplyr – Data manipulation and wrangling
- tidyr – Data reshaping and tidy data
- ggplot2 – Data visualization

## 2. Load Dataset
We will use the built-in mtcars dataset.
```{r}
data(iris)  #instead of mtcars 
head(iris)
```

## 3. Understanding the Dataset
```{r}
str(iris)
dim(iris)
summary(iris)
```

**Questions**

- How many observations are there?
- How many variables are there?
- Which variables are numerical?

## 4. Histogram
A histogram shows the distribution of a numerical variable.

Example: Miles per Gallon (mpg)
```{r}
hist(
  iris$Sepal.Length,
  main = "Distribution of Sepal Length",
  xlab = "Sepal Length",
  col = "lightblue",
  border = "black"
)

```

**Interpretation**

- Is the distribution symmetric?
- Is it positively or negatively skewed?
- Where are most observations concentrated?

### Histogram using ggplot2
```{r}
ggplot(
  iris,
  aes(x = Sepal.Length)
) +
  geom_histogram(
    bins = 12,
    fill = "steelblue",
    color = "black"
  ) +
  labs(
    title = "Distribution of Sepal Length",
    x = "Sepal Length",
    y = "Frequency"
  )
```

## 5. Boxplot
A boxplot summarizes:

- Minimum
- Q1
- Median
- Q3
- Maximum
- Outliers

```{r}
boxplot(
  iris$Sepal.Length,
  main = "Boxplot of Sepal Length",
  col = "lightgreen"
)
```

### Boxplot using ggplot2
```{r}
ggplot(
  iris,
  aes(y = Sepal.Length)
) +
  geom_boxplot(
    fill = "orange"
  ) +
  labs(
    title = "Boxplot of Sepal Length"
  )
```


## 6. Visual Identification of Outliers
Create a sample dataset with an obvious outlier.

```{r}
heights <- c(
  150, 155, 158, 160, 162,
  165, 168, 170, 172, 175,
  300
)

boxplot(
  heights,
  main = "Boxplot with Outlier",
  col = "pink"
)
```

**Discussion**

- Which value appears to be an outlier?
- Why does it appear separately?

### Variation and Distribution Spanned by Boxplots 

```{r}

# Comparing Sepal Length across different Species
ggplot(iris, aes(x = Species, y = Sepal.Length, fill = Species)) +
  geom_boxplot(alpha = 0.7, outlier.color = "#E74C3C", outlier.shape = 16) +
  scale_fill_brewer(palette = "Blues") +
  labs(
    title = "Sepal Length Across Species",
    x = "Species",
    y = "Sepal Length",
    fill = "Species Legend"
  ) +
  theme_minimal() +
  theme(legend.position = "none")
```

### Detecting Outliers Using IQR
```{r}
Q1 <- quantile(heights, 0.25)

Q3 <- quantile(heights, 0.75)

IQR_value <- IQR(heights)

lower <- Q1 - 1.5 * IQR_value

upper <- Q3 + 1.5 * IQR_value

heights[
  heights < lower |
  heights > upper
]
```

## 7. Scatterplots using ggplot2
Scatterplots help visualize relationships between variables.
```{r}
ggplot(
  iris,
  aes(
    x = Petal.Length,
    y = Petal.Width
  )
) +
  geom_point(
    size = 3,
    color = "blue"
  ) +
  labs(
    title = "Petal Length vs Petal Width",
    x = "Petal Length",
    y = "Petal Width"
  )
```


**Questions**

- Is there a positive or negative relationship?
- Is the relationship strong or weak?

## 8. Adding a Trend Line

```{r}
ggplot(
  iris,
  aes(
    x = Petal.Length,
    y = Petal.Width
  )
) +
  geom_point(size = 3) +
  geom_smooth(
    method = "lm",
    se = FALSE,
    color = "red"
  ) +
  labs(
    title = "Petal Length vs Petal Width"
  )
```

**Interpretation**

Does MPG increase or decrease as weight increases?

## 9. Correlation Analysis
**Correlation Coefficient**

```{r}
cor(
  iris$Petal.Length,
  iris$Petal.Width
)
```

**Interpretation**

| Correlation | Interpretation   |
| ----------- | ---------------- |
| +1          | Perfect positive |
| 0           | No relationship  |
| -1          | Perfect negative |


**Correlation Matrix**

```{r}
# Exclude the categorical Species column (column 5) for correlation
cor_matrix <- cor(iris[, 1:4])

cor_matrix
```

**Visualizing Correlation Matrix**

```{r}
corrplot(
  cor_matrix,
  method = "color",
  type = "upper"
)
```

**Questions**

- Which variables are strongly correlated?
- Which variables have weak correlations?

## 10. Pairwise Scatterplots
```{r}
pairs(
  iris[, 1:4],
  main = "Pairwise Scatterplots"
)
```




# ---- I used Iris instead of mtcars in above 10 examples 



## List Key Findings

1. Sepal Length: Ranges 4.5–7.9 cm; virginica is largest, setosa is smallest.

2. Outliers: The value 300 sits far past the $IQR$ boundary, confirming a severe anomaly.Petal 

3. Trend: A strong positive linear relationship links petal length and petal width.

4. Species Clustering: Setosa forms a completely isolated group due to its tiny petals.

5. Strongest Correlation: Petal length and width are highly correlated at ~0.96.

6. Weakest Correlation: Sepal width and length show almost no linear relationship.

and more

## Exercises
### Exercise 1
Perform EDA on the variable:

`mtcars$hp`

Tasks:

1. Create a histogram.
2. Create a boxplot.
3. Identify any outliers.
4. Calculate summary statistics.
5. Describe the distribution.

### Exercise 2
Using the iris dataset: `data(iris)`

Perform the following:

1. Create a histogram of Sepal Length.
2. Create a boxplot of Petal Length.
3. Create a scatterplot of Sepal Length vs Petal Length.
4. Calculate the correlation between Sepal Length and Petal Length.
5. Write a short EDA report (5–8 bullet points).

## Home work

Choose a public dataset from Kaggle or UCI and prepare a one-page EDA report containing:

- Dataset description
- Summary statistics
- Histogram
- Boxplot
- Scatterplot
- Correlation analysis
- Three key findings

